---
title: "EdX Data Science Capstone Project"
author: "William Mee"
date: '2022-06-27'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# EdX Data Science: Capstone Project

This is the machine learning capstone project for EdX Data Science online course.

I elected to implement a recommendation system based on the 
[Movielens 10 million dataset](https://grouplens.org/datasets/movielens/10m/). 

## Baseline: mean of all ratings
I followed the coursework initially to establish a baseline RMSE by taking a 
simple mean of all the ratings: this value is 1.061202. 

```{r}
rmse <- function(true_ratings, predicted_ratings) {
  sqrt(mean((true_ratings - predicted_ratings)^2))
}

mu_hat <- mean(edx$rating)
pred_0 <- rep(mu_hat, nrow(validation))
rmse(validation$rating, pred_0)
```

## Bias models (user, movies and genres)
I proceeded to calculate the bias successively of movies, users and genres. Code
snippets are below. Using this approach, I obtained the following RMSE values:

| Model | RMSE | 
|-------:|:------:|
| Movie Bias | 0.9439087 | 
| Movie/User Bias | 123 |
| 1 | 1 |  

```{r}
movie_bias <- edx %>%
  group_by(movieId) %>%
  summarize(biasMovie = mean(rating)-mu_hat)

pred_1 <- validation %>%
  left_join(movie_bias, by='movieId') %>%
  pull(biasMovie)

pred_1 <- pred_1 + mu_hat

# Prediction 1: 0.9439087
rmse(validation$rating, pred_1)

```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
